The Economics of Wildfire Aviation

### 2023-09-06 manuscript/manuscript.qmd
updated ms and added exu planes to figure 2

Over the past 10 years, the federal government has spent an average of \$2.3 billion per year on wildfire suppression [@suppress]. These costs are expected to rise as climate change drives more intense fires over longer seasons [@burke2021; @abatzoglou2016]. Approximately 21% (\$493 million) of that is spent on aviation resources [@stonesifer2021]. These costs involve hourly flight costs to operate the aircraft, the cost of retardant, and the cost to keep aircraft available and maintain it. Given the high cost of aviation, it is critical to determine whether the benefits of their application exceed the costs. Otherwise, resources could be better invested in other firefighting resources such as wildland firefighting engines, dozers, and personnel.

@@

Wildfire response is a coordinated effort between federal, state, and local land management agencies. Aviation resources have been used in wildfire response since WWII [@pyne1982] when military aircraft were retrofit with equipment that would allow them to carry and drop water. Over the years, government agencies have acquired, retrofit, and maintained a fleet of airplanes and helicopters. These aircraft have been used to drop water and retardant on or in advance of the fire, deploy smoke jumpers, to gather information about the extent and intensity of the fire, and to survey damage. While aviation can provide valuable services, they are one of the most expensive resources on an active fire.

Until 2008?, federal agencies owned and operated a fleet of aircraft bearing the cost of maintenance. Since 2008 (check this), the aircraft used by federal agencies are owned by private entities. Contracts are either exclusive use or call-when-needed where the name references the structure of the contract and the associated costs. Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days. Call-when-needed contracts specify that an aircraft is available for a short period of time (e.g., 3 days) at a daily availability rate. The optimal structure, cost, and duration of contracts is an economics problem.

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The direct cost of using aviation to fight fire is recorded because the government is responsible for paying the bill. Firefighting aviation also carries risks, which are not monetized. [@butler2015] document that 78 of the 298 (26%) recorded wildland firefighting fatalities were attributable to aviation incidents. The benefits of aviation are harder to quantify because they involve the avoided costs and losses of fires that were prevented from growing. If aviation resources reduce the probability that a fire spreads and damages property, we don't observe the counterfactual - what would have happened in the absence of aviation use. A more nuanced challenge is that aviation is not allocated randomly, but rather assigned to fires that decision maker's believe resources will have the best chance to affect fire outcomes. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

Aviation resources such as large airtankers (LAT) are believed to be most effective during the initial attack phase (the first burning period which is usually from ignition to 10:00 AM the following day). The logic is simple, aggressively attack the fire while it is small and it will not grow large. Most fires (more than 90%) are contained during initial attack. However, the fires that do escape initial containment because of weather or lack of sufficient IA effort can grow large and pose risk to values at risk.

Aviation is increasingly used to support extend attack - the incident phase after the first burning period. Indeed, [@stonesifer2016] find that in 2010 aviation was used on IA 63% of the time (see Table 3). By 2012, aviation use on IA represented only 44% of drops. A US Forest Service Aerial Firefighting Use and Effectiveness (AFUE) Report documents that between 2015 and 2018, aviation resources were used on extended attack 60% to 70% of the time (cite AFUE). Aviation resources may be used during extended attack to contain or slow fire spread, reduce burning intensity, and to reinforce fire line built on the ground (cite AFUE).

The fundamental question is when is using aviation to fight fire worth the cost. The answer depends on the objectives of the mission, which aircraft are conducting the mission, and tactically where on the fire aviation is used in relation to other suppression activities. The AFUE report provides the best available information on the effectiveness of aviation resources in accomplishing stated objectives. It finds that...

@@

Here we define a study design intended to objectively and quantitatively measure the benefits of aviation. Specifically, we outline a difference-in-differences method to assess the probability that an aviation drop stops or reduces the intensity of the fire. Since we know the time and location of the aviation drop, the fire, landscape characteristics, and weather, we can build a dataset that follows the structure of a two by two DiD design. We know the location of the drop and the direction that the fire is burning, so we can spatially define the pre and post of the treatment area. The challenge in any causal inference problem is to find an appropriate control to serve as the counterfactual. In this case, we can define a control as the area adjacent to the end of the aviation drop. The area near the end of the drop is likely to have similar vegetation and topography that influences wildfire behavior. (reference figure)

@@

We cannot draw conclusions from a single drop. We can repeat this procedure across many drops on many fires. Once the regions are defined for each drop, we can use GIS to extract information on vegetation, topography, and weather conditions at the time the fire encountered the drop. The following model can be used to estimate the average treatment effect on the treated,

@@

The identification of the treatement effect is based on the assumption that the precise location of the start and end of the drop is as good as random. The pilots intend to drop retardant in the area, but the exact location is subject to pilot accuracy, speed, altitude, and weather conditions. If the end points of the retardant drop are random, then the untreated region near the drop is just as likely to be treated.

This framework is simple but powerful. First, it produces a causal estimate of the average effectiveness of large airtanker drops toward stopping the spread of fire. Second, we can easily investigate heterogeneity of the treatment effect by interacting topography, vegetation, and weather with $(Treated_i \cdot Post_t)$. These results can help inform strategic operations highlighting when and where large airtanker drops are most effective. The results also serve as a crucial step in estimating the benefits - avoided damage - of LAT drops. For example, at time $t$, assume there is some risk to assets denominated by $\rho(e_0)*V$ where $\rho(e_0)$ is a probability of loss as a function of effort, $e_0$, and $V$ is the dollar value of the assets. If a LAT drop represents effort $e_1 > e_0$, then the benefit of the drop is $V*\left[\rho(e_0) - \rho(e_1) \right]$.

@@

The USFS determined that owning and maintaining its own fleet of aircraft is not worth the cost. In xxxx, they changed to a procurement model in which private firefighting aircraft are contracted for a period of time at a specified cost. The contacts specify a number of safety requirements, but also a daily availability price and an hourly flight price (cite Stoneseifer). The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CNW). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season? The answer depends on expected use patterns of the aviation resources.

Expected use patterns are a function of fire activity throughout the season. If the demand for aircraft was equal on every day of the season, it would be optimal to sign all exclusive use contracts for the season. If demand was intermittent but known throughout the season, it would be optimal to sign CWN contracts (up to a certain daily cost rate), so that you only pay for aviation when you need it. The reality is somewhere in between these extremes. Wildfire activity is largely driven by predictable weather patterns starting in the spring in the southwestern US, progressing north and counterclockwise across the northwest through the summer, and ending in California in the Fall. However, the fire activity on any particular day throughout the season is very unpredictable. Therefore, the optimal portfolio of contract types depends on the variability and uncertainty in demand throughout the season.

Figure 2 illustrates the variability and uncertainty of initial attack fires (bottom panel) and large fires being actively suppressed (extended attack - top panel) throughout the season. According to the 2022 USFS Schedule of Items, most large airtankers under exclusive use contracts started between March and May and were scheduled to end in September, October, or November.[^1] While the 18 large and very large airtankers under contract in 2022 span the dates when fire activity is expected to be high, they are not activated in a pattern that follows the number of IA and EA fires in figure 2. The trends in figure 2 suggest that given current usage patterns, the USFS could delay activating some LATs and VLATs until later in June. Call When Needed contracts could be used to handle surge capacity and unexpected increases in fire activity.

[^1]: The only publicly accessible document we could locate was from June 24, 2022 https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf

![Figure 2. Trends of Extended Attack (top) and Initial Attack (bottom) fires from 2007 to 2021. The points represent the median and the lines represent the interquartile range (25th to 75th percentile).](../figures/ia_ea_fires.png)

@@

-   more flexible CWN contracts: Current records suggest that CWN contracts are executed for 90 days. The federal firefighting agencies could benefit from shorter and more flexible contracts. If there were no transactions cost[^2], the ideal arrangement would allow the firefighting agency to use (and pay for) the aircraft only when they need it.

[^2]: By transactions cost, we mean the time and effort to arrange, verify, and execute the contract, the moving of aircraft, and any other logistical issues involved in using the aircraft.

@@

-   The schedule of items suggests that there are relatively few private entities that own aircraft. Procurement auctions tend to be more economically efficient when there are more bidders competing for the limited contract. Auction theory suggests that the competition between the bidders over a limited number of contracts will incentivize them to reveal their true willingness to accept, which should equal their marginal cost of provision. Moreover, the auction efficiency assumes that each bidder's cost of provision is unknown to the other bidders. When the number of bidders is small and there is ample opportunity for the bidders to communicate in advance of submitting bids, it is more likely that the bidders will bid above their marginal cost of provision. Moreover, the aircraft owners repeatedly participate in this annual procurement process allowing them to learn about their competitors and the government's willingness to pay. In principle, these forces may lead to higher procurement costs. However, these auction inefficiencies may be unavoidable since the ownership and maintenance of large aircraft is capital intensive, which serves as a barrier to new entrants in the market.[^3]

[^3]: The McDonnell Douglas DC 10, one of the largest active airtankers, is currently listed for sale for \$2.75 million (https://myemail.constantcontact.com/McDonnell-Douglas-DC-10-30-Freighter-For-Sale--Price-Reduction-.html?soid=1102673403687&aid=zTC4sV-oTNI, accessed September 9, 2023).

>>>>

Over the past 10 years, the federal government has spent an average of \\\$2.3 billion per year on wildfire suppression [@suppress]. These costs are expected to rise as climate change drives more intense fires over longer seasons [@burke2021; @abatzoglou2016]. Approximately 21% (\$493 million) of that \\\$2.3 billion is spent on aviation resources [@stonesifer2021]. These costs involve hourly flight costs to operate the aircraft, the cost of retardant, and the cost to keep aircraft available. Given the high cost of aviation, it is critical to determine whether the benefits of their application exceed the costs. Otherwise, resources could be better invested in other firefighting resources such as wildland firefighting engines, dozers, and personnel [@stonesifer2021].

@@

Wildfire response is a coordinated effort between federal, state, and local land management agencies. Firefighting resources include hand crews, engines, dozers and aircraft. These resources are owned by different agencies, but will coordinate under unified command when incidents become large and complex. Aviation resources have been used in wildfire response since WWII [@pyne1982] when military aircraft were retrofit with equipment that would allow them to carry and drop water. Over the years, government agencies have acquired, retrofit, and maintained a fleet of airplanes and helicopters. These aircraft have been used to drop water and retardant on or in advance of the fire, deploy smoke jumpers, to gather information about the extent and intensity of the fire, and to survey damage. While aviation can provide valuable services, they are one of the most expensive resources on an active fire.

Until 2008?, federal agencies owned and operated a fleet of aircraft bearing the cost of maintenance. Since 2008 (check this), the aircraft used by federal agencies are owned by private entities. Contracts are either exclusive use (EXU) or call-when-needed (CWN) where the name references the structure of the contract. Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at an agreed upon daily availability rate and hourly flight cost. Call-when-needed contracts specify that an aircraft is available for a short period of time at a daily availability rate and hourly flight cost that generally exceeds the exclusive use rates.

Something on the aircraft owners

Aircraft must be outfit with tracking technology to qualify for use by federal agencies. The Automated Telemetry Unit tracks the time-stamped location of the aircraft as well as the start and end of the drop event. These data provide the precise time and location of aviation effort; information that is crucial to quantitative analysis of the effectiveness of aviation effort.

@@

Cost benefit analysis requires the accounting of the costs and benefits of an intervention over a relevant time period [@boardman2018]. The direct cost of using aviation to fight fire is recorded because the government is responsible for payment. Firefighting aviation also carries risks, which are not monetized. [@butler2015] document that 78 of the 298 (26%) recorded wildland firefighting fatalities were attributable to aviation incidents between 2000 and 2013. The benefits of aviation are harder to quantify because they involve the avoided costs and losses of fires that were prevented from growing. If aviation resources reduce the probability that a fire spreads and damages property, we don't observe the counterfactual - what would have happened in the absence of aviation use. A more nuanced challenge is that aviation is not allocated randomly, but rather assigned to fires that decision maker's believe resources will have the best chance to affect fire outcomes. This section describes how aviation resources are currently used to fight fire, discusses the conceptual challenges to measuring benefits, and proposes data driven approaches to measuring the benefits of aviation given currently available data.

Aviation resources such as large airtankers (LAT) are believed to be most effective during the initial attack phase (the first burning period which is usually from ignition to 10:00 AM the following day). The logic is simple, aggressively attack the fire while it is small and it will not grow large. Most fires (more than 90%) are contained during initial attack. However, the fires that do escape initial containment because of weather or lack of sufficient IA effort can grow large and pose risk to values at risk.[^1]

[^1]: Values at risk is an umbrella term that captures assets that society values, including private property, infrastructure, cultural heritage sites, endangered species habitat.

Aviation is increasingly used to support extend attack - the incident phase after the first burning period. Indeed, [@stonesifer2016] find that in 2010 aviation was used on IA 63% of the time (see Table 3 in [@stonesifer2016]). By 2012, aviation use on IA represented only 44% of drops. A US Forest Service Aerial Firefighting Use and Effectiveness (AFUE) Report documents that between 2015 and 2018, aviation resources were used on extended attack 60% to 70% of the time (cite AFUE). Aviation resources may be used during extended attack to contain or slow fire spread, reduce burning intensity, and to reinforce fire line built on the ground (cite AFUE).

The fundamental question is: when is using aviation to fight fire worth the cost. The answer depends on the objectives of the mission, which aircraft are conducting the mission, and tactically where on the fire aviation is used in relation to other suppression activities. The AFUE report provides the best available information on the effectiveness of aviation resources in accomplishing stated objectives. It finds that...

@@

Here we define a study design intended to objectively and quantitatively measure the benefits of aviation. Specifically, we outline a difference-in-differences (DiD) approach to assess the probability that an aviation drop stops or reduces the intensity of the fire. Since we know the time and location of the aviation drop, the fire progression, landscape characteristics, and weather, we can build a dataset that follows the structure of a two by two DiD design. We know the location of the drop and the direction that the fire is burning, so we can spatially define the pre and post of the treatment area. The challenge in any causal inference problem is to find an appropriate control to serve as the counterfactual. In this case, we can define a control as the area adjacent to the end of the aviation drop. The area near the end of the drop is likely to have similar vegetation and topography that influences wildfire behavior. (reference figure)

@@

We cannot draw conclusions from a single drop. We repeat this procedure across many drops on many fires. Once the regions are defined for each drop, we can use GIS to extract information on vegetation, topography, and weather conditions at the time the fire encountered the drop. The following model can be used to estimate the average treatment effect on the treated,

@@

The identification of the treatement effect is based on the assumption that the precise location of the start and end of the drop is as good as random. The pilots intend to drop retardant in the area, but the exact location is subject to pilot accuracy, speed, altitude, and weather conditions. If the end points of the retardant drop are quasirandom, then the untreated region near the drop is just as likely to be treated.

This framework is simple but powerful. First, it produces a causal estimate of the average effectiveness of large airtanker drops toward stopping the spread of fire. Second, we can easily investigate heterogeneity of the treatment effect by interacting topography, vegetation, and weather with $(Treated_i \cdot Post_t)$. These results can help inform strategic operations, highlighting when and where large airtanker drops are most effective. The results also serve as a crucial step in estimating the benefits - avoided damage - of LAT drops. For example, at time $t$, assume there is some risk to assets denominated by $\rho(e_0)*V$ where $\rho(e_0)$ is a probability of loss as a function of effort, $e_0$, and $V$ is the dollar value of the assets at risk.\^\[Assets like homes generally have observable values, whereas cultural heritage sites may not. Moreover, many object to the strict use of monetary value of assets that may imply the benefits of protecting high-value homes are higher than protecting low-value homes. The principles described here accommodate alternative definitions of values at risk that may include subjective values.\] If a LAT drop represents effort $e_1 > e_0$, then the benefit of the drop is $V*\left[\rho(e_0) - \rho(e_1) \right]$. This benefit is comparable to the known cost of the flight based on the hourly cost and cost of retardant.

@@

The USFS determined that owning and maintaining its own fleet of aircraft is not worth the cost. In xxxx, they changed to a procurement model in which private firefighting aircraft are contracted for a period of time at a specified cost. The contacts specify a number of safety requirements, but also a daily availability price and an hourly flight price (cite Stoneseifer). The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CNW). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season?[^2] The answer depends on expected use patterns of the aviation resources.

[^2]: The Rand Corporation was asked to answer this question in xxxx. They concluded that the USFS should make more and better use of "scooper" aircraft that are capable of refilling water tanks by flying over water bodies. The results of this report were largely dismissed by leaders in the firefighting community for making heroic assumptions about the ability of scoopers.

Expected use patterns are a function of fire activity throughout the season. If the demand for aircraft was equal on every day of the season, it would be optimal to sign all exclusive use contracts for the season. If demand was intermittent but known throughout the season, it would be optimal to sign CWN contracts (up to a certain daily cost rate differential), so that you only pay for aviation when you need it. The reality is somewhere in between these extremes. Wildfire activity is largely driven by predictable weather patterns starting in the spring in the southwestern US, progressing north and counterclockwise across the northwest through the summer, and ending in California in the Fall. However, the fire activity on any particular day throughout the season is very unpredictable. Therefore, the optimal portfolio of contract types depends on the variability and uncertainty in demand throughout the season.

Figure 2 illustrates the variability and uncertainty of initial attack fires (bottom panel) and large fires being actively suppressed (extended attack - top panel) throughout the season. According to the 2022 USFS Schedule of Items, most large airtankers under exclusive use contracts started between March and May and were scheduled to end in September, October, or November.[^3] While the 18 large and very large airtankers under contract in 2022 span the dates when fire activity is expected to be high, they are not activated in a pattern that follows the number of IA and EA fires in figure 2. The trends in figure 2 suggest that given current usage patterns, the USFS could delay activating some LATs and VLATs until later in June. Call When Needed contracts could be used to handle surge capacity and unexpected increases in fire activity.

[^3]: The only publicly accessible document we could locate was from June 24, 2022 https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf

![Figure 2. Trends of Extended Attack (top) and Initial Attack (middle) fires from 2007 to 2021. The points represent the median and the lines represent the interquartile range (25th to 75th percentile). The bottom panel represents the number of LATs and VLATs available by day under exclusive use contract in 2022.](../figures/ia_ea_fires.png)

@@

-   more flexible CWN contracts: Current records suggest that CWN contracts are executed for 90 days. The federal firefighting agencies could benefit from shorter and more flexible contracts. If there were no transactions cost[^4], the ideal arrangement would allow the firefighting agency to use (and pay for) the aircraft only when they need it.

[^4]: By transactions cost, we mean the time and effort to arrange, verify, and execute the contract, the moving of aircraft, and any other logistical issues involved in using the aircraft.

@@

-   The schedule of items suggests that there are relatively few private entities that own aircraft. Procurement auctions tend to be more economically efficient when there are more bidders competing for the limited contract. Auction theory suggests that the competition between the bidders over a limited number of contracts will incentivize them to reveal their true willingness to accept, which should equal their marginal cost of provision. Moreover, the auction efficiency assumes that each bidder's cost of provision is unknown to the other bidders. When the number of bidders is small and there is ample opportunity for the bidders to communicate in advance of submitting bids, it is more likely that the bidders will bid above their marginal cost of provision. Moreover, the aircraft owners repeatedly participate in this annual procurement process allowing them to learn about their competitors and the government's willingness to pay. In principle, these forces may lead to higher procurement costs. However, these auction inefficiencies may be unavoidable since the ownership and maintenance of large aircraft is capital intensive, which serves as a barrier to new entrants in the market.[^5]

[^5]: The McDonnell Douglas DC 10, one of the largest active airtankers, is currently listed for sale for \$2.75 million (https://myemail.constantcontact.com/McDonnell-Douglas-DC-10-30-Freighter-For-Sale--Price-Reduction-.html?soid=1102673403687&aid=zTC4sV-oTNI, accessed September 9, 2023).
